---
title: "Practicum 2"
author:
- Peter Claussen
- Maggie Germundson
- Stephanie Liebl
date: "10/26/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(cluster)
library(ggdendro)
```

# Introduction

We have been give a large data file, consisting of expression levels of RNA for different genes. The mRNA were obtained from multiple cell lines. The client requests that we use RNA expression patterns to group cell lines. These data are expected to be very sparse.

## Data Dimensions

```{r,echo=FALSE}
if(!file.exists('single.cell.Rda')) {
  cell.dat <- read.csv('single_cell.csv',header=TRUE)
  save(cell.dat,file='single.cell.Rda')
} else {
  load(file='single.cell.Rda')
}
#first column are the gene identifiers
genes <- cell.dat[,1]
#first row are barcodes for cell lines
bar.code <- cell.dat[1,]
cell.dat <- cell.dat[,-1]
dim(cell.dat)
```

What is the variation in maximum expression levels?

```{r}
row.max <- apply(cell.dat,1,max)
hist(row.max)
```


These data represent approximately 13000 genes and 2700 cell lines. For simplicity at this stage, we'll assume all 2700 cell lines are valid. 

We could potentially use all 13000+ gene expression levels to generate a cluster analysis for cell lines. This, however, is computationally intractable, so we wish to reduce the dimensions of the problem. Briefly, we consider four dimension reduction procedures.

```{r,echo=FALSE,eval=FALSE}
distances <- dist(t(cell.dat), method="euclidean")
clusters <- hclust(distances,method="ward.D")
ggdendrogram(clusters,rotate=TRUE)
```

We consider the expression pattern of some randomly selected genes.

```{r,fig.width=6,fig.height=8}
random.samples <- sample(1:dim(cell.dat)[1],12)
random.samples
par(mfrow=c(4,3))
for(i in random.samples) {
  tmp <- unlist(cell.dat[i,])
  hist(tmp,main=paste("Histogram,",genes[i]))
}
```


## Commonly expressed genes

For many genes, it appears that only a small proportion of cell express these genes. We may use this as a dimension reduction procedure. We can, for example, exclude all genes where the median expression value is 0 (that is, genes that are not expressed in at least one-half of all cell lines.). 

```{r}
row.median <- apply(cell.dat,1,median)
dup <- row.median
summary(row.median)
dup <- row.median[row.median>0]
hist(dup)
work.dat <- cell.dat[row.median>0,]
genes.common <- genes[row.median>0]
dim(work.dat)
```

We plot the histograms of some of the common genes

```{r,fig.width=6,fig.height=8}
print(common.sample <- sample(genes.common,12))
common.mask <- genes %in% common.sample
common.dat <- cell.dat[common.mask,]
common.names <- genes[common.mask]
par(mfrow=c(4,3))
for(i in 1:dim(common.dat)[1]) {
  tmp <- unlist(common.dat[i,])
  hist(tmp,main=paste("Histogram, ",common.names[i]))
}
```



This leaves us with 273 genes, a more reasonable number to work with. A proposed dendrogram, showing potential cell clusters follows.

```{r,fig.width=6,fig.height=8}
distances <- dist(t(work.dat), method="euclidean")
clusters <- hclust(distances,method="ward.D")
ggdendrogram(clusters,rotate=TRUE)
```

```{r,fig.width=6,fig.height=8}
heatmap(t(work.dat))
```


## Most commonly expressed genes

We might further reduce the number by selecting those with gene expression patterns approximately normal. Since there are still a large number, we use an approximation. When data are normally distributed (or, at least, symmetrically distributed), then the median will be close to the mean. For the current subset of genes, we can visualize this by plotting the median gene expression level against the corresponding mean. We bound this by a band representing 95-105% - that is, we wish to select genes where the mean is within 5%, plus or minus, of the mean.

```{r,echo=FALSE}
row.means <- apply(cell.dat,1,mean)
row.median <- apply(cell.dat,1,median)
plot(row.median,row.means)
abline(0,1,col='blue')
abline(0,1.05,col='red')
abline(0,0.95,col='red')
```

```{r}
ratio <- row.means/row.median
mask <- ratio<1.05 & ratio>0.95
sum(mask)
work.dat <- cell.dat[mask,]
genes.work <- genes[mask]
dim(work.dat)
```

We print some sample histograms for common genes with approximately symmetric distributions.

```{r,fig.width=6,fig.height=8}
print(work.sample <- sample(genes.work,12))

work.mask <- genes %in% work.sample
common.dat <- cell.dat[work.mask,]
common.names <- genes[work.mask]
par(mfrow=c(4,3))
for(i in 1:dim(common.dat)[1]) {
  tmp <- unlist(common.dat[i,])
  hist(tmp,main=paste("Histogram, ",common.names[i]))
}
```

A potential dendrogram based on this subset of genes follows.

```{r,fig.width=6,fig.height=8}
distances <- dist(t(work.dat), method="euclidean")
clusters <- hclust(distances,method="ward.D")
ggdendrogram(clusters,rotate=TRUE)
```


```{r,fig.width=6,fig.height=8}
heatmap(t(work.dat))
```

## Uncommon, but not rare, genes

Distinct cell types may be distiquished not simply by gene expression levels, but by gene expression patterns. That is, a certain gene may only be expressed in certain cell types and could thus be usefully diagnostic, while other genes (i.e. glycolytic pathway) are expressed in nearly all cells and may be of little use in determining cell groups. 

Since we have seen that the majority genes are expressed in less than half the cells, we don't wish to use very rare genes. Instead, we propose to consider genes that are expressed in at least 10% of cell lines, but no more than 20%.

```{r}
incidence.dat <- cell.dat
incidence.dat[incidence.dat>0.01] <- 1

incidences <- apply(incidence.dat,1,mean)
hist(incidences)

incidence.dat <- incidence.dat[incidences>0.1 & incidences<0.2,]
genes.incidence <- genes[incidences>0.1 & incidences<0.2]
dim(incidence.dat)
```

```{r,fig.width=6,fig.height=8}
print(incidence.sample <- sample(genes.incidence,12))

incidence.mask <- genes %in% incidence.sample
tmp.dat <- cell.dat[incidence.mask,]
tmp.names <- genes[incidence.mask]
par(mfrow=c(4,3))
for(i in 1:dim(tmp.dat)[1]) {
  tmp <- unlist(tmp.dat[i,])
  hist(tmp,main=paste("Histogram, ",tmp.names[i]))
}
```

This results in approximately 1100 candidate genes. A potential dendrogram is presented below:

```{r,fig.width=6,fig.height=8}
distances <- dist(t(incidence.dat), method="euclidean")
clusters <- hclust(distances,method="ward.D")
ggdendrogram(clusters,rotate=TRUE)
```

```{r,fig.width=6,fig.height=8}
heatmap(t(incidence.dat))
```

## 1) Gene Ontology

We may be able to reduce the number of genes considered with further guidance from the researcher. A brief survey of the descriptors associate with rows suggest families of genes may be used to reduce the dimensions of the problem. For example, there appear to be 12 different aldehyde dehydrogenase genes represented in these data. 


```{r}
sum(grepl('^ALDH',genes))
```

It also appears these data include non-protein coding genes, for example, we find examples of Long Intergenic Non-Protein Coding RNA (LINC). We also find non-cytoplasmic proteins such as Mitochondrial Ribosomal Protein (MRPL)

```{r}
sum(grepl('^LINC',genes))
sum(grepl('^MRPL',genes))
```

Other examples of proteins with known functionality include membrane protiens, i.e.

```{r}
#Solute Carrier
sum(grepl('^SLC',genes))
#Transmembrane Protein
sum(grepl('^TMEM',genes))
```

Other groups include proteins with identifiable secondary or tertiary structural characteristics,
```{r}
#coiled coil domain containing
sum(grepl('^CCDC',genes))
#Leucine-rich repeat-containing protein
sum(grepl('^LRRC',genes))
#WD Repeat Domain
sum(grepl('^WDR',genes))
#Zinc Finger Protein
sum(grepl('^ZNF',genes))
#Family with sequence similarity
sum(grepl('^FAM',genes))
```

while some genes have not clear ontology (at least, based on naming conventions)

```{r}
#open reading frame
sum(grepl('orf',genes))
sum(grepl('A[C|L|P][0-9]',genes))
sum(grepl('^CT[B|C]',genes))
sum(grepl('^RP[1-9]',genes))
```

### Question 1.

Is there are subset of genes that, *a priori* would be preferred to differentiate among cell lineages? For example, perhaps zinc-finger protein expression is discriminatory?

```{r,fig.width=6,fig.height=8}
znf.genes <- genes[grepl('^ZNF',genes)]
znf.dat <- cell.dat[grepl('^ZNF',genes),]
distances <- dist(t(znf.dat), method="euclidean")
clusters <- hclust(distances,method="ward.D")
ggdendrogram(clusters,rotate=TRUE)
```

```{r,fig.width=6,fig.height=8}
print(znf.sample <- sample(znf.genes,12))

znf.mask <- genes %in% znf.sample
tmp.dat <- cell.dat[znf.mask,]
tmp.names <- genes[znf.mask]
par(mfrow=c(4,3))
for(i in 1:dim(tmp.dat)[1]) {
  tmp <- unlist(tmp.dat[i,])
  hist(tmp,main=paste("Histogram, ",tmp.names[i]))
}
```


```{r,fig.width=6,fig.height=8}
heatmap(t(znf.dat))
```

## PCA 

A final approach to dimension reduction is to use principal component analysis to identify pseudovariables that capture variation in gene expression patterns. We present a PCA of the most commonly expressed genes identified in the previous step. PCA may be computationally intractable for the full data set.

```{r}
cell.pca <- prcomp(t(work.dat), center = TRUE,scale. = TRUE)
#summary(cell.pca)
```

A biplot suggests the two component may be discriminatory, but further components may not be discriminatory.
```{r}
#library(devtools)
#install_github("vqv/ggbiplot", force = TRUE)

library(ggbiplot)
ggbiplot(cell.pca)
ggbiplot(cell.pca,choices=c(2,3))
```



```{r,eval=FALSE}
col.median <- apply(cell.dat,2,median)
col.mean <- apply(cell.dat,2,mean)
dup <- col.median
hist(col.mean)
summary(col.median)
summary(col.mean)
keep.cols <- col.mean>0.2
work.dat <- cell.dat[,keep.cols]
dim(work.dat)
work.dat <- work.dat[,col.median>1]
bar.code <- bar.code[col.median>1]
```
