---
title: "Practicum 1"
author: "Peter Claussen"
date: "9/16/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

We have been given an Excel data file with data collected for a study from a call center. These data include health metric from **employees** gathered over an eight month period. 

Variables that are of particular interest have been highlighted in the Excel file. These include variables relating to weight change. We have been tasked to provide an analysis to answer two questions:

- Does total metabolic minutes have an effect on weight gain?
- Does shift have an effect on weight gain?

# Data Cleaning

There are many missing rows in the data. While complete data screening will not be determined until we've chosen both response and predictor variables, we can perform some prelimiries.

We have inspected the data in Excel format manually, and have exported to CSV to read into R for analysis. For convenience, we have identified the columns of interest. 

Two orange highlighted columns ('lbs_gained', 'pounds_gained') appear to be identical. We will include only `lbs_gained`

```{r}
yellow <- c('shift','Total_Met_Min')
orange <- c('gender', 'Age', 'height', 'weightgain', 'lbs_gained', 'BMI', 'Vig.ex.Time', 'Mod.ex.time', 'Walk.ex.Time')
```

Also from inspection, the column `Snumber` appears to uniquely identify observations, and when this column is empty, the remaining columns are also empty. Thus, we will use this to in a first pass to screen for missing data.

```{r}
metrics.dat <- read.csv('Practicum 1 Data.csv',header=TRUE)
metrics.dat <- metrics.dat[!is.na(metrics.dat$Snumber),]
```

```{r}
summary(metrics.dat[,c(yellow,orange)])
```

There are two responses, gender and weightgain, that should be binomial. However, there are three levels.

```{r}
metrics.dat$gender[metrics.dat$gender==''] <- 'Missing'
metrics.dat$weightgain[metrics.dat$weightgain==''] <- 'Missing'
unique(metrics.dat$gender)
unique(metrics.dat$weightgain)
```

```{r}
metrics.dat$shift <- factor(metrics.dat$shift)
metrics.dat$gender <- factor(metrics.dat$gender)
metrics.dat$weightgain <- factor(metrics.dat$weightgain)
```

# Response Variables
While the data include a variable (data column) named `weightgain`, we are asked to consider other reponse variables, including change in weight (`lbs_gained`) and change in BMI. Thus, our first task is to determine the appropriate response variable. The choice of response variable will dictate both choice of statistical method (i.e. logistic regression vs linear regression) and methods for data cleaning.

## weightgain (Binomial Response)

A quick plot to visualize

```{r}
metrics.dat$Total_Met_Min[is.na(metrics.dat$Total_Met_Min)] <- 0
plot(weightgain ~ Total_Met_Min,data=metrics.dat)
plot(weightgain ~ shift,data=metrics.dat)
```

At first glance, this appears to be a logistic regression - the response variable (weightgain) should be binary (Yes/No). However, we are asked to consider alternate response variables (BMI, pounds gained) as markers for the general response "weight gain', as opposed to the specific data variable 'weightgain'

## Pounds gained (Continuous Response)

## Change in BMI

BMI at the end of the study period has been recorded, while initial BMI has not. We can, however, calculate an initial BMI from other data columns.

"Beginning BMI is unknown. The beginning weight can be calculated by body weight at 8 months –pounds gained. Then this value along with the height variable can be used to calculate beginning BMI; not that this calculation differs whether using kg/meters or pounds/inches" 


```{r, echo=FALSE,results='hide'}
# Calculate body weight at 8 months
metrics.dat['beginning_bweight'] <- metrics.dat$bweight - metrics.dat$pounds_gained

# Need to account for instances where body weight is not entered
metrics.dat$beginning_bweight[metrics.dat$beginning_bweight<='0'] <- '0'

# Take a look at all BMI values
unique(metrics.dat$BMI)

# Calculate beginning BMI -- internet tells me BMI=(weight/height^2) * 703
# Since height values ~60/65, I am assuming they are inches not meters
metrics.dat['beginning_bweight'] <- as.numeric(metrics.dat$beginning_bweight)
metrics.dat['beginning_BMI'] <- (metrics.dat$beginning_bweight / (metrics.dat$height)^2)*703

# ignore non-positive values
metrics.dat$beginning_BMI[is.nan(metrics.dat$beginning_BMI)] <- 0
metrics.dat$beginning_BMI[!is.finite(metrics.dat$beginning_BMI)] <- 0

# Take a look at all beginning_BMI values
unique(metrics.dat$beginning_BMI)

```


```{r, echo=FALSE, results='hide'}
##Look at BMI in the beginning vs at the end

# convert to numeric for calculations
metrics.dat$beginning_BMI <- as.numeric(metrics.dat$beginning_BMI)

# Are the means different? --Yes, BMI is bigger than beginning_BMI
mean(metrics.dat$beginning_BMI)
mean(metrics.dat$BMI)
```


From the histograms of the employees' beginning BMI and the employees' final BMI, we see that there is a slight shift to the right in the frequencies in the histogram for Final BMI. This suggests to us that there were higher numbers of higher BMI among the employees. If BMI is used as our response variable, we will look greater at the relationship between beginning BMI and final BMI, as well as the relationship between shift and BMI, and total metabolic minutes and BMI, respectively. 
```{r, echo=TRUE}
# Visualize the columns
par(mfrow=c(1,2))

metrics.dat$beginning_BMI[metrics.dat$beginning_BMI==0] <- NA
hist(as.numeric(metrics.dat$beginning_BMI),main='Beginning BMI', xlab='Beginning BMI')

metrics.dat$BMI[metrics.dat$BMI==0] <- NA
hist(as.numeric(metrics.dat$BMI), main='Final BMI', xlab='Final BMI')
```

```{r, echo=FALSE, results='hide'}
# reset par() 
dev.off()
```

In the scatter plot visualizing (`beginning BMI`) versus final BMI ('`BMI`'), it appears that there is a strong linear relationship between the variables. The scatter plot also suggests that (`metrics.dat$BMI`) is frequently greater than (`metrics.dat$beginning_BMI`). 

```{r, echo=TRUE}
# Look at relationship between BMI and beginning BMI -- appears that BMI only increases or stays the same over time
plot(metrics.dat$beginning_BMI, metrics.dat$BMI, xlab='Beginning BMI', ylab='Final BMI')

```

The following histograms aim to see if there is any apparent relationship between gender and BMI gain in this dataset. It appears there is a difference in histogram shapes for female beginning BMI and female final BMI. If BMI is chosen to be the response variable, these variables will be looked at more closely to determine if there is a positive/negative relationship.
```{r, echo=TRUE}
# To view all plots together
par(mfrow=c(2,2))

# Female Beginning BMI
hist((metrics.dat[which(metrics.dat$gender=='Female'), 'beginning_BMI']), main="Female Beginning BMI", xlab='Gender is Female')

# Male Beginning BMI
hist((metrics.dat[which(metrics.dat$gender=='Male'), 'beginning_BMI']), main="Male Beginning BMI", xlab='Gender is Male')

# Female BMI
hist((metrics.dat[which(metrics.dat$gender=='Female'), 'BMI']), main="Female BMI", xlab='Gender is Female')

# Male BMI
hist((metrics.dat[which(metrics.dat$gender=='Male'), 'BMI']), main="Male BMI", xlab='Gender is Male')


```

```{r, echo=FALSE, results='hide'}
# return to default par() options
dev.off()
```
 
To look at the difference between the initial BMI and the final BMI for call center employees, we created a new column (`BMI_diff`) by subtracting the initial BMI from the final BMI. Thus, in this column, positive values represent increases in BMI over the eight month period.
```{r, echo=FALSE}
# Create a BMI_diff column to show BMI change over the 8 months
metrics.dat['BMI_diff'] <- metrics.dat$BMI - metrics.dat$beginning_BMI

max(metrics.dat$BMI_diff)
min(metrics.dat$BMI_diff)

# need to look at this without the zeros that are from not having BMI values
plot(metrics.dat$BMI_diff)
```

```{r, echo=FALSE, results='hide'}
# Were the BMI calculations correct?
# BMI = (weight/height^2)*703
temp <- ''
temp <- ((metrics.dat$bweight)/(metrics.dat$height)^2)*703

plot(temp, metrics.dat$BMI)
abline(0,1,col='red')
```

Observe plots of (`shift`) and (`CalcTMM`) vs (`BMI_diff`)
```{r}
par(mfrow=c(1,2))
plot(metrics.dat$shift, metrics.dat$BMI_diff, xlab='Shift', ylab='Difference in BMI')

plot(metrics.dat$CalcTMM, metrics.dat$BMI_diff, xlab='CalcTMM', ylab='Difference in BMI')
```

Observe plot of (`Gender`) vs (`BMI_Diff`)
```{r}
plot(metrics.dat$gender,metrics.dat$BMI_diff)
```


# Predictor Variables

Total Metabolic Minutes is calculate from 3 data columns; this variable is missing for many rows. We are given the formula

```
Total_met_min = 8*Vig_ex_time + 4*Mod_ex_time + 3.3*Walk_ex_time
```


```{r}
metrics.dat$CalcTMM <- with(metrics.dat, 8*Vig.ex.Time + 4*Mod.ex.time + 3.3*Walk.ex.Time)

plot(CalcTMM ~ Total_Met_Min,data=metrics.dat)
abline(0,1,col='red')
plot(weightgain ~ CalcTMM,data=metrics.dat)
```


```{r}
plot(lbs_gained ~ CalcTMM,data=metrics.dat)
```

```{r}
plot(lbs_gained ~ weightgain,data=metrics.dat)
```


```{r,fig.width=10,fig.height=10}
pairs(metrics.dat[,c(yellow,orange)])
```

This summary doesn't include missing values. To visualize, let's set NA to 0

```{r}
metrics.dat[is.na(metrics.dat)] <- 0
```

```{r,fig.width=10,fig.height=10}
pairs(metrics.dat[,c(yellow,orange)])
```



