---
title: "Practicum 1"
author:
- Peter Claussen
- Ben Derenge
- Stephanie Leibl
date: "9/16/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,echo=FALSE}
metrics.dat <- read.csv('Practicum 1 Data.csv',header=TRUE)
metrics.dat <- metrics.dat[!is.na(metrics.dat$Snumber),]
```

# Overview

We have been given an Excel data file with data collected for a study from a call center. These data include health metrics from **employees** gathered over an eight month period. 

Variables that are of particular interest have been highlighted in the original data. These include variables relating to weight change and demographics. Specifically, we will consider in this preliminary proposal the variables

```{r,echo=FALSE}
#define constants to be used as column selection arguments
yellow <- c('shift','Total_Met_Min')
orange <- c('gender', 'Age', 'height', 'weightgain', 'lbs_gained', 'BMI', 'Vig.ex.Time', 'Mod.ex.time', 'Walk.ex.Time')
```


### Highlighted in yellow

- `shift`
- `Total_Met_Min`

### Highlighted in orange
- `gender`
- `Age`
- `height`
- `weightgain`
- `lbs_gained`
- `BMI`
- `Vig.ex.Time`
- `Mod.ex.time`
- `Walk.ex.Time`

We will use the convention that text rendered in `sans serif` font denote variable or column names found in the original data set, or derived variables calculated from columns in the original data, while *italics* will denote real-world processes or phenomena of interest. Thus, `weightgain` denotes the data column in the original data file, while *weight gain* denotes some measure of the change in weight by individuals over the study period.


. We have been tasked to provide an analysis to address two specific aims:

- **(SA1)** Does *total metabolic minutes* have an effect on *weight gain*?
- **(SA2)** Does *shift* have an effect on *weight gain*?

We will refer to as **SA1** and **SA2** in further discussion. We address *total metabolic minutes*, *weight gain* and *shift* in the following sections.


# Total MET minutes

*Total MET Minutes* is a composite measure obtained from survey responses (International Physical Activity Questionnaire (IPAQ) short form) to estimate overall physical activity. Physical activity is divided into categories described as *vigorous*, *moderate* and *walking*.  
Total Metabolic Minutes can be calculated from 3 data columns. We are given the formula
```
Total_met_min = 8*Vig_ex_time + 4*Mod_ex_time + 3.3*Walk_ex_time
```
to calculate *Total MET Minutes* from the data.


`Total_met_min` contains many missing values. We will create a new data column, `CalcTMM` that is calclulated from existing data columns as given in the formula above. To visualize the number of missing values, we plot missing values as 0 in the following figure.

```{r,fig.cap="Total MET Minutes calculated from data, plotted against the data column title Total_Met_Min",echo=FALSE}
metrics.dat$Total_Met_Min[is.na(metrics.dat$Total_Met_Min)] <- 0
metrics.dat$CalcTMM <- with(metrics.dat, 8*Vig.ex.Time + 4*Mod.ex.time + 3.3*Walk.ex.Time)

plot(CalcTMM ~ Total_Met_Min,data=metrics.dat)
abline(0,1,col='red')
```

We can see that the original data column `Total_met_min` has a large number of missing values. If we use this data column as given, we will reduce the number of observations in the analysis. This graph also confirms that the formula given above was used to compute `Total_met_min` from the data.

# Shift

```{r,echo=FALSE, include=FALSE}
#determine 
unique(metrics.dat$shift)
```

*Shift* takes values of the form `7am`, `8am`, `...`, `2pm`, `other`. There are some columns that have missing values for `shift`. We propose that $shift$ be modeled as an ordinal data type, with missing values grouped with `other` and `other` takes an ordinal value greater than `2pm`.

```{r}
metrics.dat$shift[metrics.dat$shift==''] <- 'other'
shift.levels <- c(paste(c(7:11),'am',sep=''),paste(c(12,1:2),'pm',sep=''),'other')
metrics.dat$shift <- factor(metrics.dat$shift,shift.levels)
summary(metrics.dat$shift)
```


# Weight Gain

While the data include a variable (data column) named `weightgain`, we are asked to consider other reponse variables, including change in weight (`lbs_gained`) and change in BMI. Thus, our first task is to determine the appropriate response variable. The choice of response variable will dictate both choice of statistical method (i.e. logistic regression vs linear regression) and methods for data cleaning.

## weightgain (Binomial Response)

`weightgain` in the original data contains missing values. 

```{r,echo=FALSE}
metrics.dat$weightgain <- factor(metrics.dat$weightgain)
```


```{r, fig.width=12,fig.height=6,fig.cap="`weightgain` versus Total MET Minutes (calculated) and Shift"}
par(mfrow=c(1,2))
plot(weightgain ~ CalcTMM,data=metrics.dat)
plot(weightgain ~ shift,data=metrics.dat)
```

At first glance, this appears to be a logistic regression - the response variable (weightgain) should be binary (Yes/No). However, we are asked to consider alternate response variables (BMI, pounds gained) as markers for the general response "weight gain', as opposed to the specific data variable 'weightgain'

## Pounds gained (Continuous Response)
The 'pounds_gained' columns contains only nonnegative integer values and missing values. This implies every known observation gained weight. Checking corresponding values in the binary `weightgain` column reveals all the NA in `pounds_gaine`' are "No" in `weightgain`. This will affect the interpretation of SA.1 and SA.2. The client will need to specify if we concerned about only magnitude of weight gained, and completely ignore weight lost. Or, analyze the factors associated with a binary outcome of 'any weight lost' vs 'any weight gained'. 


### Excluding 120 missing from `pounds_gained`
Taking the first approach, we would only be concerned with observation that gained weight, excluding 120 missing rows and 1 zero.
```{r}
g <- subset(metrics.dat, metrics.dat$pounds_gained >0)
par(mfrow=c(1,2))
hist(g$pounds_gained,breaks=30)
hist(log(g$pounds_gained),breaks=30)
```
The column `pounds_gained` can be log transformed, to adjusted its skew.

```{r, echo=FALSE}
#shift vs lbs gained
#$ts <- as.factor((g$shift))
#plot(g$ts,log(g$pounds_gained), ylab="pounds_gained", xlab="shift")
#plot(g$shift,g$shift)
#plot(as.factor(g$shift),log(g$pounds_gained), ylab="pounds_gained", ###xlab="shift")

#table(g$shift)
s <- ordered(g$shift, levels=c("7am","8am","9am","10am","11am","12pm","1pm","2pm","Missing","other"))
#boxplot(pounds_gained ~ s, data=g)
par(mfrow=c(1,2))
boxplot(pounds_gained ~ s, data=g, xlab="shift")

plot(g$CalcTMM, g$pounds_gained)
#plot(g$CalcTMM, log(g$pounds_gained))
```
The bar plot above suggest shift does have potential to predict pounds_gained. The column 'pounds_gained' seems to decrease on average as shifts move later into the day, with an exception for the 2pm shift, which looks closest to  7am and not like 1pm. Missing shifts appear to have gained the most weight of all the shifts. This plot supports that notion that shifts could be ranked as opposed to treating shifts like independent factors.


### Targeting Binary outcome with `weightgain`
```{r}
boxplot(metrics.dat$Total_Met_Min,as.factor(metrics.dat$weightgain), ylab="Total Met Minutes", xlab="weightgain")
```
If desired, `weightgain` can be predicted from Total_met_minutes, as most of the "No"s from `weighgain` have NAs or zeros from `total_met_minutes`. This might change if we predict the missing TMM values using the eqaution from earlier.






## Change in BMI

BMI at the end of the study period has been recorded, while initial BMI has not. We can, however, calculate an initial BMI from other data columns.

"Beginning BMI is unknown. The beginning weight can be calculated by body weight at 8 months –pounds gained. Then this value along with the height variable can be used to calculate beginning BMI; not that this calculation differs whether using kg/meters or pounds/inches"
```{r}
# Calculate body weight at 8 months
metrics.dat['beginning_bweight'] <- metrics.dat$bweight - metrics.dat$pounds_gained

# Need to account for instances where body weight is not entered
metrics.dat$beginning_bweight[metrics.dat$beginning_bweight<='0'] <- '0'

# Take a look at all BMI values
unique(metrics.dat$BMI)

# Calculate beginning BMI -- internet tells me BMI=(weight/height^2) * 703
# Since height values ~60/65, I am assuming they are inches not meters
metrics.dat['beginning_bweight'] <- as.numeric(metrics.dat$beginning_bweight)
metrics.dat['beginning_BMI'] <- (metrics.dat$beginning_bweight / (metrics.dat$height)^2)*703

# Take a look at all beginning_BMI values
unique(metrics.dat$beginning_BMI)

# ignore non-positive values
metrics.dat$beginning_BMI[is.nan(metrics.dat$beginning_BMI)] <- 0
metrics.dat$beginning_BMI[!is.finite(metrics.dat$beginning_BMI)] <- 0
```

Look at BMI in the beginning vs at the end
```{r}
# convert to numeric for calculations
metrics.dat$beginning_BMI <- as.numeric(metrics.dat$beginning_BMI)

# Are the means different? --Yes, BMI is bigger than beginning_BMI
mean(metrics.dat$beginning_BMI)
mean(metrics.dat$BMI)

# Visualize the columns
metrics.dat$beginning_BMI[metrics.dat$beginning_BMI==0] <- NA
hist(as.numeric(metrics.dat$beginning_BMI))

metrics.dat$BMI[metrics.dat$BMI==0] <- NA
hist(as.numeric(metrics.dat$BMI))

# Look at relationship between BMI and beginning BMI -- appears that BMI only increases or stays the same over time
plot(metrics.dat$beginning_BMI, metrics.dat$BMI)

```

I want to see if gender affects BMI 
```{r}
# To view all plots together
par(mfrow=c(2,2))

# Female Beginning BMI
hist(log(metrics.dat[which(metrics.dat$gender=='Female'), 'beginning_BMI']), main="Log of Female Beginning BMI", ylim=c(0,40))

# Female BMI
hist(log(metrics.dat[which(metrics.dat$gender=='Female'), 'BMI']), main="Log of Female BMI", ylim=c(0,40))

# Male Beginning BMI
hist(log(metrics.dat[which(metrics.dat$gender=='Male'), 'beginning_BMI']), main="Log of Male Beginning BMI", ylim=c(0,40))

# Male BMI
hist(log(metrics.dat[which(metrics.dat$gender=='Male'), 'BMI']), main="Log of Male BMI", ylim=c(0,40))

```







# Data Review

There are many missing rows in the data. While complete data screening will not be determined until we've chosen both response and predictor variables, we can perform some prelimiries.

We have inspected the data in Excel format manually, and have exported to CSV to read into R for analysis. For convenience, we have identified the columns of interest. 

Two orange highlighted columns ('lbs_gained', 'pounds_gained') appear to be identical. We will include only `lbs_gained`

Also from inspection, the column `Snumber` appears to uniquely identify observations, and when this column is empty, the remaining columns are also empty. Thus, we will use this to in a first pass to screen for missing data.

```{r}
summary(metrics.dat[,c(yellow,orange)])
```

There are two responses, gender and weightgain, that should be binomial. However, there are three levels.

```{r}
metrics.dat$gender[metrics.dat$gender==''] <- 'Missing'
metrics.dat$weightgain[metrics.dat$weightgain==''] <- 'Missing'
unique(metrics.dat$gender)
unique(metrics.dat$weightgain)
```

```{r}
metrics.dat$shift <- factor(metrics.dat$shift)
metrics.dat$gender <- factor(metrics.dat$gender)
metrics.dat$weightgain <- factor(metrics.dat$weightgain)
```




# Other Predictor Variables



**TODO**

```{r}
metrics.dat[is.na(metrics.dat)] <- 0
```

```{r,fig.width=10,fig.height=10}
pairs(metrics.dat[,c(yellow,orange)])
```



# Points for consideration

## Selection of response variable. 

We have proposed three response variables of interest as surrogates for weight gain, specifically `weightgain`, `pounds_gained`,and @@@@.

### Pounds Gained
Pounds gained is skewed data, so if this is our response variable, we would consider either data transformation, or a generalized linear model of the poisson family. 

Also, there are 121 fewer responses to use, as those observations are missing.
The box-whisker plot `pounds_gained` vs `shift` shows a curvolinear response. There should exist a straightforward model predicting weight_gained from shift. The relationship between weight_gained and Calculated `Total Met Minutes` would require further exploratory analysis.

## Additional predictor variables.

**SA1** and **SA2** include only one predictor variable each (`Total_Met_Min` and `shift`, respectively). This implies a simple regression of one-way ANOVA analysis. However, the data includes other demographic variables that may be confounded with the primary ("yellow") predictors

- Does the client wish us to perform mutliple regression and variable selection with the "orange" highlighted variables?
- Does the client wish us to perform multiple regression and variable selection with any additional variables in the original data file?

## Scope of analysis

The questions of interest only reference two variables, while there are multiple highlighted data columns. The first suggests a single predictor variable, but it would be best to perform multiple regression

