---
title: "Practicum 1"
author: "Peter Claussen"
date: "9/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

We have been given an Excel data file with data collected for a study from a call center. These data include health metric from **employees** gathered over an eight month period. 

Variables that are of particular interest have been highlighted in the Excel file. These include variables relating to weight change. We have been tasked to provide an analysis to answer two questions:

- Does total metabolic minutes have an effect on weight gain?
- Does shift have an effect on weight gain?

# Data Cleaning

There are many missing rows in the data. While complete data screening will not be determined until we've chosen both response and predictor variables, we can perform some prelimiries.

We have inspected the data in Excel format manually, and have exported to CSV to read into R for analysis. For convenience, we have identified the columns of interest. 

Two orange highlighted columns ('lbs_gained', 'pounds_gained') appear to be identical. We will include only `lbs_gained`

```{r}
yellow <- c('shift','Total_Met_Min')
orange <- c('gender', 'Age', 'height', 'weightgain', 'lbs_gained', 'BMI', 'Vig.ex.Time', 'Mod.ex.time', 'Walk.ex.Time')
```

Also from inspection, the column `Snumber` appears to uniquely identify observations, and when this column is empty, the remaining columns are also empty. Thus, we will use this to in a first pass to screen for missing data.

```{r}
metrics.dat <- read.csv('Practicum 1 Data.csv',header=TRUE)
metrics.dat <- metrics.dat[!is.na(metrics.dat$Snumber),]
```

```{r}
summary(metrics.dat[,c(yellow,orange)])
```

There are two responses, gender and weightgain, that should be binomial. However, there are three levels.

```{r}
metrics.dat$gender[metrics.dat$gender==''] <- 'Missing'
metrics.dat$weightgain[metrics.dat$weightgain==''] <- 'Missing'
unique(metrics.dat$gender)
unique(metrics.dat$weightgain)
```

```{r}
metrics.dat$shift <- factor(metrics.dat$shift)
metrics.dat$gender <- factor(metrics.dat$gender)
metrics.dat$weightgain <- factor(metrics.dat$weightgain)
```

# Response Variables
While the data include a variable (data column) named `weightgain`, we are asked to consider other reponse variables, including change in weight (`lbs_gained`) and change in BMI. Thus, our first task is to determine the appropriate response variable. The choice of response variable will dictate both choice of statistical method (i.e. logistic regression vs linear regression) and methods for data cleaning.

## weightgain (Binomial Response)

A quick plot to visualize

```{r}
metrics.dat$Total_Met_Min[is.na(metrics.dat$Total_Met_Min)] <- 0
plot(weightgain ~ Total_Met_Min,data=metrics.dat)
plot(weightgain ~ shift,data=metrics.dat)
```

At first glance, this appears to be a logistic regression - the response variable (weightgain) should be binary (Yes/No). However, we are asked to consider alternate response variables (BMI, pounds gained) as markers for the general response "weight gain', as opposed to the specific data variable 'weightgain'

## Pounds gained (Continuous Response)

## Change in BMI

BMI at the end of the study period has been recorded, while initial BMI has not. We can, however, calculate an initial BMI from other data columns.


# Predictor Variables

Total Metabolic Minutes is calculate from 3 data columns; this variable is missing for many rows. We are given the formula

```
Total_met_min = 8*Vig_ex_time + 4*Mod_ex_time + 3.3*Walk_ex_time
```


```{r}
metrics.dat$CalcTMM <- with(metrics.dat, 8*Vig.ex.Time + 4*Mod.ex.time + 3.3*Walk.ex.Time)

plot(CalcTMM ~ Total_Met_Min,data=metrics.dat)
abline(0,1,col='red')
plot(weightgain ~ CalcTMM,data=metrics.dat)
```


```{r}
plot(lbs_gained ~ CalcTMM,data=metrics.dat)
```

```{r}
plot(lbs_gained ~ weightgain,data=metrics.dat)
```


```{r,fig.width=10,fig.height=10}
pairs(metrics.dat[,c(yellow,orange)])
```

This summary doesn't include missing values. To visualize, let's set NA to 0

```{r}
metrics.dat[is.na(metrics.dat)] <- 0
```

```{r,fig.width=10,fig.height=10}
pairs(metrics.dat[,c(yellow,orange)])
```



